~\Large Increased sexual dimorphism evolves in a fossil stickleback following ecological release from fish piscivores

### 2023-08-08 manuscript/paper.tex
Moved imputed data set to data folder, and updated STAN code appropriately to run faster.
Update manuscript

\title{~\Large Increased sexual dimorphism evolves in a fossil
stickleback following ecological release from fish piscivores}
\author{\large Allison Ozark\(^1\), Akhil Ghosh\(^1\), Raheyma
Siddiqui\(^1\), Matthew Stuart\(^1\),\\
\large  Samantha Swank\(^1\), Michael A. Bell\(^2\), Gregory J.
Matthews\(^1\), and Yoel E. Stuart\(^1\) \vspace{-1.1mm}\\
\strut \\
\large \(^1\) Loyola University Chicago, Chicago, IL, USA
\vspace{-1.1mm}\\
\strut \\
\large \(^2\) University of California Museum of Paleontology, Berkeley,
CA, USA \vspace{-1.1mm}\\}
\date{}

@@

Ecological release theory suggests that a prey population's niche should
expand when predation is relaxed or removed (reviewed in Herrmann,
Stroud, and Losos (\protect\hyperlink{ref-Herrmann2021}{2021})). Niche
changes in habitat and resource use should then drive phenotypic
evolution as the population adapts to new conditions (Herrmann, Stroud,
and Losos (\protect\hyperlink{ref-Herrmann2021}{2021})). For example,
resource use expansion should result in trophic trait evolution (cite;
e.g.,). Habitat expansion should result in locomotor change (cite;
e.g.,). And release from enemies should result in reduction of costly
defense traits (cite; e.g.,).

Niche expansion could manifest through a combination of increasing
within- and between-individual niche widths (Bolnick et al.
(\protect\hyperlink{ref-Bolnick2010}{2010}); Herrmann, Stroud, and Losos
(\protect\hyperlink{ref-Herrmann2021}{2021})), including divergence
between the sexes (Bolnick and Doebeli
(\protect\hyperlink{ref-Bolnick2003}{2003}); Cooper, Gilman, and
Boughman (\protect\hyperlink{ref-Cooper2011}{2011})). Theory and
empirical data suggest that a population presented with novel ecological
opportunity might experience disruptive selection on males and females
stemming from intraspecific competition over newly accessible resources,
resulting in intersexual divergence in habitat use and associated
phenotypes (Schoener 1968; Shine
(\protect\hyperlink{ref-Shine1989}{1989}); Bolnick and Doebeli
(\protect\hyperlink{ref-Bolnick2003}{2003}); Butler, Sawyer, and Losos
(\protect\hyperlink{ref-Butler2007}{2007}); Bolnick and Lau
(\protect\hyperlink{ref-Bolnick2008}{2008}); Cooper, Gilman, and
Boughman (\protect\hyperlink{ref-Cooper2011}{2011}); but see Stuart et
al. (\protect\hyperlink{ref-Stuart2021}{2021}); Blain
(\protect\hyperlink{ref-Blain2022}{2022})). Such character displacement
between the sexes is therefore one explanation for sexual dimorphism (de
Lisle papers).

We tested this link between release from predators and the evolution of
sexual dimorphism in a well-preserved, high resolution lineage of the
fossil stickleback fish (Gasterosteus doryssus). The sex of the
individual fossils, which is unobserved in practice, was imputed using
multiple imputation by chained equations (MICE) (Buuren and
Groothuis-Oudshoorn (\protect\hyperlink{ref-MICE}{2011})) with observed
data used to train the imputation model based on extant stickleback with
known sex. Then we tracked the multivariate evolution of sexual
dimorphism over \textasciitilde16,000 years for traits related to
swimming, feeding and defense. The fossil sequence captures a bout of
adaptive evolution following ecological release from predators, as
follows. The fossil lineage appeared in the depositional environment as
a fully armored form, with complete pelvic girdles, two pelvic spines,
and three dorsal spines, on average (Michael A. Bell, Travis, and Blouw
(\protect\hyperlink{ref-Bell2006}{2006}); Stuart, Travis, and Bell
(\protect\hyperlink{ref-Stuart2020}{2020})). However, this habitat
appears to have been missing fish piscivores (e.g., trout and other
salmonids known to prey on modern stickleback; M. A. Bell
(\protect\hyperlink{ref-Bell2009}{2009})), relaxing putative selection
for armor (Reimchen papers, Bell papers, Bowne
(\protect\hyperlink{ref-Bowne1994}{1994}); Roesti et al.~2023). The
lineage thus adaptively evolved armor loss and reduction in several
other traits as well (Hunt, Bell, and Travis
(\protect\hyperlink{ref-Hunt2008}{2008}); Stuart, Travis, and Bell
(\protect\hyperlink{ref-Stuart2020}{2020}); Siddiqui et al.~in prep),
and tooth wear data suggest that the population also began eating more
planktonic prey, expanding to an open water lifestyle (Purnell et
al.~2007).

Sexual Dimorphism is interesting because\ldots\ldots{}

Sexual Dimoprhism lit review: Saitta et al.
(\protect\hyperlink{ref-SaittaEtAl2020}{2020})

The remainder of this manuscript contains a description of the data in
section \ref{sec:data} and a description of our models in
\ref{sec:models}. Section \ref{sec:results} presents a summary of our
results, and we end with our conclusion and future work in section
\ref{sec:conclusions}.

@@

A major challenge in this work is inferring the sex of fossil specimens.
This typically cannot be done directly, except for lineages whose sexes
are distinguished by the presence or absence of sex-specific characters
(Hone and Mallon (\protect\hyperlink{ref-Hone2017}{2017}); Saitta et al.
(\protect\hyperlink{ref-SaittaEtAl2020}{2020})). Instead,
paleobiologists resort to statistical detection of sex and sexual
dimorphism, including tests for bimodality in trait distributions (e.g.,
Hone and Mallon (\protect\hyperlink{ref-Hone2017}{2017})) and divergence
in growth curves (e.g., Saitta et al.
(\protect\hyperlink{ref-SaittaEtAl2020}{2020})).

We used a third approach to infer fossil sex and study the evolution of
sexual dimorphism by taking advantage of the fact that G. doryssus is
part of the extant threespine stickleback species complex (Gastersteus
aculeatus). Along with the fossil data collected for this study, we also
have data from extant populations of the Threespine Stickleback,
Gasterosteus aculeatus, measuring the same traits for individuals whose
sex could be determined directly. We then combined these sets of data
and treat the sex of the fossil specimens as missing and used multiple
imputation (Little and Rubin
(\protect\hyperlink{ref-little2002statistical}{2002})) to impute the sex
of the fossils many times. Imputations were performed using the MICE
(Buuren and Groothuis-Oudshoorn (\protect\hyperlink{ref-MICE}{2011}))
algorithm and implemented in R (R Core Team
(\protect\hyperlink{ref-R2022language}{2022})) with M = 100 imputations
performed.

After imputing the sex of the fossil data, each completed data set is
then used to fit a modified Ornstein--Uhlenbeck (OU) (Uhlenbeck and
Ornstein (\protect\hyperlink{ref-OUProcess}{1930})) model using a
Bayesian framework to look for sexual dimorphism across a variety of
stickleback phenotypes

@@

The fossil data consists of 814 specimens with unknown sex over 18 time
periods spaced \textasciitilde1000 years apart. Table 2 shows the sample
size at each of the 18 time periods. There are at least 22 specimens at
each time period with a high of 67 specimens in period 7.

\begin{longtable}[]{@{}
  >{\raggedright\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.3492}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.1905}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.4603}}@{}}
\caption{Traits and trait descriptions. `sc' denotes size correction of
trait against standard length. Names of bones follow Bowne
(\protect\hyperlink{ref-Bowne1994}{1994}) unless otherwise
noted.}\tabularnewline
\toprule()
\begin{minipage}[b]{\linewidth}\raggedright
Trait Name
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Trait Code
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Trait Description
\end{minipage} \\
\midrule()
\endfirsthead
\toprule()
\begin{minipage}[b]{\linewidth}\raggedright
Trait Name
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Trait Code
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Trait Description
\end{minipage} \\
\midrule()
\endhead
Standard Length & sl & Distance from anterior tip of premaxilla to
posterior end of last vertebra (hypural plate) \\
Dorsal Spine & mds & Number of dorsal spines from 0 to 3 \\
Dorsal Fin Ray & mdf & Number of bones in the dorsal fin posterior to
the third dorsal spine (i.e., soft dorsal fin rays) \\
Anal Fin Ray & maf & Number of bones in the anal fin posterior to the
anal spine (i.e., soft dorsal fin rays) \\
Abdominal Vertebra & mav & Number of vertebrae anterior to the first
vertebra contacting an anal fin pterygiophore (Aguirre et al.~2014) \\
Caudal Vertebra & mcv & Number of vertebrae posterior to and including
the first vertebra contacting an anal fin pterygiophore (Aguirre et
al.~2014) \\
Pterygiophore number & mpt & Number of pterygiophores anterior to but
excluding the pterygiophore under the third dorsal spine, which is
immediately anterior to and contiguous with the dorsal fin \\
Pelvic Spine length & lps.sc & Length from the base of one pelvic spine
above its articulation with the pelvic girdle to its distal tip \\
Ectocoracoid & ect.sc & Length between the anterior and posterior tips
of the shoulder girdle base (i.e., ectocoracoid) \\
Pelvic Girdle & tpg.sc & Length between the anterior to posterior tips
along midline. If vestigial, the sum of longest anterior-posterior axis
for the vestiges \\
Cleithrum length & cle.sc & Length from free dorsal tip to ventral tip
of the cleithrum on the anterior margin of the shoulder girdle (i.e.,
cleithrum) \\
Premaxilla & pmx.sc & Length from the anterior tip of the premaxilla to
the distal tip of the ascending process of the premaxilla \\
Dorsal Spine & Ds\#.sc\# = 1,2,or 3 & Length from the base of a dorsal
spine above the pterygiophore to its distal tip along the anterior
edge \\
Pterygiophore & lpt.sc & Distance between the anterior to posterior tips
of the pterygiophore immediately preceding the 3rd dorsal spine (when
present) \\
\bottomrule()
\end{longtable}

\begin{table}[ht]
\centering
\caption{The number of stickleback fossils at each time point.  Sample size at each time point ranges from a low of 22 to a high of 67.}
\begin{tabular}{rl}
  \hline
time & count \\ 
  \hline
  1 & 43 \\ 
    2 & 41 \\ 
    3 & 51 \\ 
    4 & 41 \\ 
    5 & 46 \\ 
    6 & 48 \\ 
    7 & 67 \\ 
    8 & 55 \\ 
    9 & 42 \\ 
   10 & 33 \\ 
   11 & 37 \\ 
   12 & 22 \\ 
   13 & 41 \\ 
   14 & 43 \\ 
   15 & 46 \\ 
   16 & 47 \\ 
   17 & 56 \\ 
   18 & 55 \\ 
   \hline
\end{tabular}
\label{tabn}

\end{table}

@@

Let \(\boldsymbol{W}\) be an \((n_{extant} + n_{fossil}) \times 1\)
vector of the covariate gender of the stickleback fish,
\(\boldsymbol{X}\) be an \((n_{extant} + n_{fossil}) \times K\) matrix
of the \(K\) continuous phenotypes of interest, and \(\boldsymbol{Y}\)
be an \((n_{extant} + n_{fossil}) \times L\) matrix of the \(L\)
discrete phenotypes of interest. Because the gender of the fossilized
stickleback fish is unobservable, we further define
\(\boldsymbol{W} = (\boldsymbol{W}_{extant}^T,\boldsymbol{W}_{fossil}^T)^T\)
where \(\boldsymbol{W}_{extant}\) and \(\boldsymbol{W}_{fossil}\) are
the \(n_{extant} \times 1\) and \(n_{fossil} \times 1\) vectors of the
observed extant gender and missing fossil gender, respectively.

We impute the missing gender for the fossil data by sampling from the
posterior predictive distribution
\(P(\boldsymbol{W}_{fossil}|\boldsymbol{W}_{extant}, \boldsymbol{X}, \boldsymbol{Y})\)
using the multiple imputation by chained equations (MICE) algorithm
(Buuren and Groothuis-Oudshoorn (\protect\hyperlink{ref-MICE}{2011}))
with predictive mean matching. Traditionally, the choice for the number
of completed dats sets is a relatively small number such as \(M = 5\) or
\(M = 10\). However, Zhou and Reiter
(\protect\hyperlink{ref-ZhouReiter2009}{2010}) recommend a larger number
of imputed data sets if the data users intend on performing Bayesians
analysis after imputation, which in this case, we do. Therefore, the
imputation algorithm is run to obtain a total of \(M = 100\) completed
datasets. In addition to this, Zhou and Reiter
(\protect\hyperlink{ref-ZhouReiter2009}{2010}) suggests rather than
using Rubin's combining rules to combine across imputed data sets,
instead pool all of the draws from the posterior distributions across
all of the imputed data sets to estimate the posterior distributions of
parameters of interest. We proceed with our Bayesian analysis in this
manner.

@@

\textcolor{red}{We should note somewhere that we are only using the $W_{fossil}$ in the modeling part.  We drop the $W_{extant}$.  So just note that $W_{ti}$ is really $W_{fossil,ti}$.  Not sure how to say this, but we need to make it clear that we are only using the fossil data for the OU modeling.   }

\hypertarget{continuous-phenotypes}{%
\subsubsection{Continuous Phenotypes}\label{continuous-phenotypes}}

For a given imputed dataset, let \(W_{ti}\) be the imputed gender,
\(\boldsymbol{X}_{ti}\) be the \(K \times 1\) vector of continuous
phenotypes, and \(\boldsymbol{Y}_{ti}\) be the \(L \times 1\) vector of
discrete phenotypes for \(t = 1, \ldots, T\) and \(i = 1\ldots,n_{t}\)).

Define the last element of \(\boldsymbol{X}_{ti}\) (\(X_{K,ti}\)) as the
length of the stickleback. For \(k = 1,\cdots,K - 1\), we will assume
\begin{align}
{X}_{k,ti} & \overset{iid}{\sim}\left\{\begin{array}{llll} \mathcal{N}(\mu_{k,ft} & + \beta_k(X_{K,ti} - \mu_{K,ft}),&\sigma_k^2), & W_{ti} = \text{Female} \\ \mathcal{N}(\mu_{k,mt} & + \beta_k(X_{K,ti} - \mu_{K,mt}),&\sigma_k^2), & W_{ti} = \text{Male} \end{array}\right. \\
{X}_{K,ti} & \overset{iid}{\sim}\left\{\begin{array}{lll} \mathcal{N}(\mu_{K,ft},&\sigma_{K}^2), & W_{ti} = \text{Female} \\ \mathcal{N}(\mu_{K,mt},&\sigma_{K}^2), & W_{ti} = \text{Male} \end{array}\right.
\label{eq:X}
\end{align} where \(\mu_{k,ft}\) and \(\mu_{k,mt}\) represent the
time-\(t\) specific mean of phenotype \(k\) for female and male
stickleback fish, respectively, and \(\beta_k\) is an additional
parameter to account for the biological phenomenon that other continuous
phenotypes for a particular animal are positively correlated with their
length \textbf{CITATION NEEDED}. We further set \begin{align}
\mu_{k,gt} = \theta_{k,g} + u_{k,gt},
\label{eq:mu}
\end{align} for \(g \in \{f,m\}\) where \(\theta_{k,g}\) is the overall
mean of phenotype \(X_k\) for each gender, and \(u_{k,gt}\) measures the
difference between \(\mu_{k,gt}\) and \(\theta_{k,g}\). We then fit
\(\boldsymbol{u}_{k,g} = \{u_{k,g1},\cdots,u_{k,gT}\}\) to an
Ornstein-Uhlenback (OU) process (Uhlenbeck and Ornstein
(\protect\hyperlink{ref-OUProcess}{1930})) where we assume \(u_{k,gT}\)
have a marginal mean of 0. Because each time period is
\textasciitilde1000 years, we will discretize the OU process without
loss of generality. For \(t = 2,\ldots,T\), we assume \begin{align}
u_{k,gt} \overset{iid}{\sim}\mathcal{N}(\kappa_{k} u_{k,g(t-1)} , \tau_k^2).
\label{eq:u_ar1}
\end{align}

\(\kappa_{k,g}\) represents the correlation between \(\mu_{k,gt}\) and
\(\mu_{k,g(t+1)}\), and it also follows that
\(\text{cor}(u_{g,t},u_{g,t+h}) = \kappa^h\) for \(h \in \mathbb{N}\).
In addition, we assume \begin{align}
u_{k,g1} \overset{iid}{\sim}\mathcal{N}\left(0,\frac{\tau_k^2}{1-\kappa_{k}^2}\right).
\label{eq:u1}
\end{align} This model choice is to preserve stationarity;
i.e.~\(p(u_{k,gs}) = p(u_{k,gt})\) for \(s \neq t\).

\hypertarget{discrete-phenotypes}{%
\subsubsection{Discrete Phenotypes}\label{discrete-phenotypes}}

>>>>

\title{~\Large Increased sexual dimorphism evolves in a fossil
stickleback following ecological release from fish piscivores}
\author{\large Allison Ozark\(^{1,*}\), Matthew Stuart\(^{2,3,*}\),
Raheyma Siddiqui\(^1\), Akhil Ghosh\(^2\)\\
\large Samantha Swank\(^1\), Michael A. Bell\(^4\), Gregory J.
Matthews\(^{2,3}\), and Yoel E. Stuart\(^{1,+}\)\\
\vspace{-1.1mm}\\
\large \(^1\) Department of Biology, Loyola University Chicago, Chicago,
IL, USA \vspace{-1.1mm}\\
\large \(^2\) Department of Mathematics and Statistics, Loyola
University Chicago, Chicago, IL, USA \vspace{-1.1mm}\\
\large \(^3\) Center for Data Science and Consulting, Loyola University
Chicago, Chicago, IL, USA \vspace{-1.1mm}\\
\large \(^4\) University of California Museum of Paleontology, Berkeley,
CA, USA \vspace{-1.1mm}\\
\large \(^*\) Equal contribution \vspace{-1.1mm}\\
\large \(^+\) Corresponding:
\href{mailto:ystuart@luc.edu}{\nolinkurl{ystuart@luc.edu}}
\vspace{-1.1mm}}
\date{}

@@

Ecological release theory suggests that a prey population's niche should
expand when predation is relaxed or removed, as prey take advantage of
new enemy free space (reviewed in Herrmann, Stroud, and Losos
(\protect\hyperlink{ref-Herrmann2021}{2021})). Niche expansion could
manifest through a combination of increasing within- and
between-individual niche widths (Bolnick et al.
(\protect\hyperlink{ref-Bolnick2010}{2010}); Herrmann, Stroud, and Losos
(\protect\hyperlink{ref-Herrmann2021}{2021})), including divergence
between the sexes (Bolnick and Doebeli
(\protect\hyperlink{ref-Bolnick2003}{2003}); Cooper, Gilman, and
Boughman (\protect\hyperlink{ref-Cooper2011}{2011})). Theory and
empirical data suggest that a population presented with ecological
opportunity following ecological release might experience disruptive
selection on males and females stemming from intraspecific competition
over newly accessible resources, resulting in intersexual divergence in
habitat use (W. (\protect\hyperlink{ref-Schoener1968}{1968}); Shine
(\protect\hyperlink{ref-Shine1989}{1989}); Bolnick and Doebeli
(\protect\hyperlink{ref-Bolnick2003}{2003}); Butler, Sawyer, and Losos
(\protect\hyperlink{ref-Butler2007}{2007}); Bolnick and Lau
(\protect\hyperlink{ref-Bolnick2008}{2008}); Cooper, Gilman, and
Boughman (\protect\hyperlink{ref-Cooper2011}{2011}); but see Stuart et
al. (\protect\hyperlink{ref-Stuart2021}{2021}); Blain
(\protect\hyperlink{ref-Blain2022}{2022})). Accordingly, sexes should
also diverge in morphological traits associated with habitat use; such
ecologically-driven ``character displacement between the sexes'' is
therefore one explanation for phenotypic sexual dimorphism (S. P. De
Lisle and Rowe (\protect\hyperlink{ref-deLisleRowe2015}{2015}); S. Paiva
De Lisle S. P. and Rowe (\protect\hyperlink{ref-deLisleNco2018}{2018})).

We tested this link between release from predators and the evolution of
sexual dimorphism in a well-preserved, finely-resolved,
\textasciitilde16,000 year-long sequence of the fossil stickleback fish
(Gasterosteus doryssus). The fossil lineage appeared in the depositional
environment as a fully armored form, with complete pelvic girdles, two
pelvic spines, and three dorsal spines, on average (Michael A. Bell,
Travis, and Blouw (\protect\hyperlink{ref-Bell2006}{2006}); Stuart,
Travis, and Bell (\protect\hyperlink{ref-Stuart2020}{2020})). However,
this habitat appears to have been missing fish piscivores (e.g., trout
and other salmonids known to prey on modern stickleback; M. A. Bell
(\protect\hyperlink{ref-Bell2009}{2009})), relaxing putative selection
for armor (T. E. Reimchen (\protect\hyperlink{ref-Reimchen1980}{1980}),
T. E. Reimchen (\protect\hyperlink{ref-Reimchen1994}{1994}), O. Bell M.
A. and Koenings (\protect\hyperlink{ref-BellEtAl1993}{1993}); Bowne
(\protect\hyperlink{ref-Bowne1994}{1994}); Roesti and Schluter
(\protect\hyperlink{ref-Roestietal2023}{2023})). The lineage adaptively
evolved armor loss but also reduction in the means of several other
traits (Hunt, Bell, and Travis (\protect\hyperlink{ref-Hunt2008}{2008});
Stuart, Travis, and Bell (\protect\hyperlink{ref-Stuart2020}{2020});
Siddiqui et al.~in prep). In particular, tooth wear data suggest that
the population shifted to eat more planktonic prey through time,
indicative of increased open water habitat use (Purnell et al.~2007). If
this divergence in habitat use is partitioned between sexes (Reimchen T.
E. and Nosil (\protect\hyperlink{ref-ReimchenNosil2004}{2004});
Spoljaric and Reimchen
(\protect\hyperlink{ref-SpoljaricReimchen2008}{2008});D. S. Reimchen T.
E. (\protect\hyperlink{ref-Reimchen2016}{2016})), then we should expect
sexual dimorphism to evolve for traits related to habitat use in G.
doryssus.

To test this prediction, we imputed the sex of individual fossils, which
is unobserved in practice, with Multiple Imputation by Chained Equations
(MICE) (Buuren and Groothuis-Oudshoorn
(\protect\hyperlink{ref-MICE}{2011})), using observed data from extant
stickleback of known sex to train the imputation model. Then we tracked
the multivariate evolution of sexual dimorphism over
\textasciitilde16,000 years for 16 traits related to swimming, feeding
and defense.

@@

A major challenge in this work is inferring the sex of fossil
stickleback specimens. Indeed, for most species, sex cannot be detected
directly, except for lineages whose sexes are distinguished by the
presence or absence of sex-specific characters (Hone and Mallon
(\protect\hyperlink{ref-Hone2017}{2017}); Saitta et al.
(\protect\hyperlink{ref-SaittaEtAl2020}{2020})). Instead,
paleobiologists resort to statistical detection of sex and sexual
dimorphism, including tests for bimodality in trait distributions (e.g.,
Hone and Mallon (\protect\hyperlink{ref-Hone2017}{2017})) and divergence
in growth curves (e.g., Saitta et al.
(\protect\hyperlink{ref-SaittaEtAl2020}{2020})). We used a third
approach to infer fossil sex and study the evolution of sexual
dimorphism by taking advantage of the fact that G. doryssus is part of
the extant threespine stickleback species complex, Gasterosteus
aculeatus. We measured the same trait set from a phenotypically diverse
set of extant G. aculeatus populations, for whom individual sex was
known. We then combined fossil and extant data sets, treated fossil sex
as missing data, and used the MICE (Buuren and Groothuis-Oudshoorn
(\protect\hyperlink{ref-MICE}{2011})) multiple imputation (Little and
Rubin (\protect\hyperlink{ref-little2002statistical}{2002})) algorithm,
and implemented in R (Team
(\protect\hyperlink{ref-R2022language}{2022})), to impute individual sex
of the fossil data set 100 times. We then fit a modified
Ornstein--Uhlenbeck (OU) (Uhlenbeck and Ornstein
(\protect\hyperlink{ref-OUProcess}{1930})) model using a Bayesian
framework to the data to test for the evolution of sexual dimorphism in
each trait.

@@

The fossil data consists of 814 specimens with unknown sex over 18 time
periods spaced \textasciitilde1000 years apart. Figure 1 shows the
sample size at each of the 18 time periods. There are at least 22
specimens at each time period with a high of 67 specimens in period 7.

\begin{figure}

{\centering \includegraphics{paper_files/figure-latex/unnamed-chunk-2-1}

}

\caption{Fossil sampel size for each time period}\label{fig:unnamed-chunk-2}
\end{figure}

\begin{longtable}[]{@{}
  >{\raggedright\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.3492}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.1905}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.4603}}@{}}
\caption{Traits and trait descriptions. `sc' denotes size correction of
trait against standard length. Names of bones follow Bowne
(\protect\hyperlink{ref-Bowne1994}{1994}) unless otherwise
noted.}\tabularnewline
\toprule()
\begin{minipage}[b]{\linewidth}\raggedright
Trait Name
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Trait Code
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Trait Description
\end{minipage} \\
\midrule()
\endfirsthead
\toprule()
\begin{minipage}[b]{\linewidth}\raggedright
Trait Name
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Trait Code
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Trait Description
\end{minipage} \\
\midrule()
\endhead
Standard Length & stl & Distance from anterior tip of premaxilla to
posterior end of last vertebra (hypural plate) \\
Dorsal Spine & mds & Number of dorsal spines from 0 to 3 \\
Dorsal Fin Ray & mdf & Number of bones in the dorsal fin posterior to
the third dorsal spine (i.e., soft dorsal fin rays) \\
Anal Fin Ray & maf & Number of bones in the anal fin posterior to the
anal spine (i.e., soft dorsal fin rays) \\
Abdominal Vertebra & mav & Number of vertebrae anterior to the first
vertebra contacting an anal fin pterygiophore (Aguirre et al.~2014) \\
Caudal Vertebra & mcv & Number of vertebrae posterior to and including
the first vertebra contacting an anal fin pterygiophore (Aguirre et
al.~2014) \\
Pterygiophore number & mpt & Number of pterygiophores anterior to but
excluding the pterygiophore under the third dorsal spine, which is
immediately anterior to and contiguous with the dorsal fin \\
Pelvic Spine length & lps.sc & Length from the base of one pelvic spine
above its articulation with the pelvic girdle to its distal tip \\
Ectocoracoid & ect.sc & Length between the anterior and posterior tips
of the shoulder girdle base (i.e., ectocoracoid) \\
Pelvic Girdle & tpg.sc & Length between the anterior to posterior tips
along midline. If vestigial, the sum of longest anterior-posterior axis
for the vestiges \\
Cleithrum length & cle.sc & Length from free dorsal tip to ventral tip
of the cleithrum on the anterior margin of the shoulder girdle (i.e.,
cleithrum) \\
Premaxilla & pmx.sc & Length from the anterior tip of the premaxilla to
the distal tip of the ascending process of the premaxilla \\
Dorsal Spine & Ds\#.sc\# = 1,2,or 3 & Length from the base of a dorsal
spine above the pterygiophore to its distal tip along the anterior
edge \\
Pterygiophore & lpt.sc & Distance between the anterior to posterior tips
of the pterygiophore immediately preceding the 3rd dorsal spine (when
present) \\
\bottomrule()
\end{longtable}

@@

Let \(\boldsymbol{W}\) be an \((n_{extant} + n_{fossil}) \times 1\)
vector of the covariate gender of the stickleback fish and
\(\boldsymbol{Y}\) be an \((n_{extant} + n_{fossil}) \times K\) matrix
of the \(K\) phenotypes of interest. Because the gender of the
fossilized stickleback fish is unobservable, we further define
\(\boldsymbol{W} = (\boldsymbol{W}_{extant}^T,\boldsymbol{W}_{fossil}^T)^T\)
where \(\boldsymbol{W}_{extant}\) and \(\boldsymbol{W}_{fossil}\) are
the \(n_{extant} \times 1\) and \(n_{fossil} \times 1\) vectors of the
observed extant gender and missing fossil gender, respectively.

We impute the missing gender for the fossil data by sampling from the
posterior predictive distribution
\(P(\boldsymbol{W}_{fossil}|\boldsymbol{W}_{extant},\boldsymbol{Y})\)
using the multiple imputation by chained equations (MICE) algorithm
(Buuren and Groothuis-Oudshoorn (\protect\hyperlink{ref-MICE}{2011}))
with predictive mean matching. Traditionally, the choice for the number
of completed dats sets is a relatively small number such as \(M = 5\) or
\(M = 10\). However, Zhou and Reiter
(\protect\hyperlink{ref-ZhouReiter2009}{2010}) recommend a larger number
of imputed data sets if the data users intend on performing Bayesians
analysis after imputation, which in this case, we do. Therefore, the
imputation algorithm is run to obtain a total of \(M = 100\) completed
datasets. In addition to this, Zhou and Reiter
(\protect\hyperlink{ref-ZhouReiter2009}{2010}) suggests rather than
using Rubin's combining rules to combine across imputed data sets,
instead pool all of the draws from the posterior distributions across
all of the imputed data sets to estimate the posterior distributions of
parameters of interest. We proceed with our Bayesian analysis in this
manner.

@@

\textcolor{red}{We should note somewhere that we are only using the $W_{fossil}$ in the modeling part.  We drop the $W_{extant}$.  So just note that $W_{ti}$ is really $W_{fossil,ti}$.  Not sure how to say this, but we need to make it clear that we are only using the fossil data for the OU modeling.  Greg, check and see if this makes sense.}

For a given imputed dataset, let \(W_{ij}\) be the imputed gender and
\(\boldsymbol{Y}_{ij}\) be the \(K \times 1\) vector of phenotypes for
stickleback fossil \(j\) at time \(t_i\) where \(i = 1, \ldots, T\) and
\(j = 1,\ldots,n_{t}\). Note, we are inputting only the fossil data into
our model; connecting to the previous section, \(W_{ij}\) and
\(\boldsymbol{Y}_{ij}\) can be interpretted as \(W_{fossil,ij}\) and
\(\boldsymbol{Y}_{fossil,ij}\). In addition, we denote \(Y_{K,ij}\), the
last variable in \(\boldsymbol{Y}_{ij}\), to be the standard length of
the fish, \begin{align}
{Y}_{K,ij} & \overset{iid}{\sim}\left\{\begin{array}{lll} \mathcal{N}(\mu_{K,ft_i},&\sigma_{K}^2), & W_{ij} = \text{Female} \\ \mathcal{N}(\mu_{K,mt_i},&\sigma_{K}^2), & W_{ij} = \text{Male} \end{array}\right..
\label{eq:stl}
\end{align}

It is reasonable to assume that the other continuous traits of
stickleback fish will have some correlation with its standard length
\textbf{(CITATION)}. We account for this by adding an additional
parameter, \(\gamma_k\), into our model. More specifically, if
\(Y_{k,ij}\) is a continuous trait, then \begin{align}
{Y}_{k,ij} & \overset{iid}{\sim}\left\{\begin{array}{llll} \mathcal{N}(\mu_{k,ft_i} + \gamma_kY_{K,ij},\sigma_k^2), & W_{ij} = \text{Female} \\ \mathcal{N}(\mu_{k,mt_i} + \gamma_kY_{K,ij},\sigma_k^2), & W_{ij} = \text{Male} \end{array}\right..
\label{eq:cont}
\end{align}

In addition, we assume that the discrete traits abdominal vertebrae
(mav) and caudal vertebrae (mcv) also have correlation with the standard
length of the fish (\textbf{CITATION}). If \(Y_{k,ij}\) is one of the
above traits, then \begin{align}
{Y}_{k,ij} & \sim \left\{\begin{array}{ll} Poisson(\exp\{\mu_{k,ft_i} + \gamma_kY_{K,ij}\}), & W_{ij} = \text{Female} \\ Poisson(\exp\{\mu_{k,mt_i} + \gamma_kY_{K,ij}\}), & W_{ij} = \text{Male} \end{array}\right..
\label{eq:disc_corr}
\end{align}

For the other discrete traits, we also assume the above model except we
set \(\gamma_k = 0\) because of the assumption of no correlation between
the standard length and these traits. More specifically, we assume
\begin{align}
{Y}_{k,ij} & \sim \left\{\begin{array}{ll} Poisson(\exp\{\mu_{k,ft_i}\}), & W_{ij} = \text{Female} \\ Poisson(\exp\{\mu_{k,mt_i}\}), & W_{ij} = \text{Male} \end{array}\right..
\label{eq:disc_ind}
\end{align}

In the above model descriptions, \(\mu_{k,ft_i}\) and \(\mu_{k,mt_i}\)
model the time-\(t_i\) specific mean of phenotype \(k\) for female and
male stickleback fish, respectively. We point out that, for the discrete
traits, the means are represented by \(\exp\{\mu_{k,ft_i}\}\) and
\(\exp\{\mu_{k,mt_i}\}\) for ease of use in our modeling technique. We
further set \begin{align}
\mu_{k,gt_i} = \beta_{0,kg} + \beta_{1,kg}t_i + u_{k,gt_i},
\label{eq:mu}
\end{align} for \(g \in \{f,m\}\) where \(\beta_{0,kg}\) and
\(\beta_{1,kg}\) are regression parameters of phenotype \(Y_k\) for each
gender, accounting for the possibility of a time-dependent trend in the
mean structure, and \(u_{k,gt_i}\) is the corresponding residual. To
account for potential correlations between the residuals for a given
trait \(k\) and gender \(g\), we fit an Ornstein-Uhlenback (OU) process
(Uhlenbeck and Ornstein (\protect\hyperlink{ref-OUProcess}{1930})). More
specifically, define \(du_{k,gt} = u_{k,g(t + dt)} - u_{k,gt}\), the
change in \(u_k,gt\) for a given trait \(k\) and gender \(g\) over a
miniscule time period \(dt\). The OU process is defined as \begin{align}
du_{k,gt} = -\kappa_k u_{k,gt} dt + \tau_k dW_t,
\label{eq:OU_cont}
\end{align} where \(\kappa_k\) is a parameter associated with the
correlation between \(u_{k,gt}\) and \(u_{k,g(t+dt)}\), \(\tau_k\) is
the standard deviation of the OU process, and \(W_t\) is a standard
Brownian motion. As shown in (Uhlenbeck and Ornstein
(\protect\hyperlink{ref-OUProcess}{1930})), the closed form solution for
the SDE in (\ref{eq:OU_cont}) is \begin{align}
u_{k,gt_i} \overset{iid}{\sim}\mathcal{N}\left(u_{k,gt_{i-1}}\exp\{-\kappa_k(t_{i} - t_{i-1})\} , \frac{\tau_k^2(1 - \exp\{-2\kappa_k(t_{i} - t_{i-1}))\}}{2\kappa_k}\right)
\label{eq:OU_sol}
\end{align} for \(i = 2,\cdots,T\). In a traditional OU process, the
initial value \(u_{k,gt_1}\) is assumed to be a (potentially unknown)
constant, and we discuss the procedure for estimating this value below.

In our empirical study, we analyze the following four nested models for
the mean process outlined in (\ref{eq:mu}):
